A finite-state morphological transducer for Kyrgyz

### 2012-03-13 paper/paper.tex
updaté

The analyser is designed to be used with the longest-match left-to-right (LRLM) tokenisation strategy, as described in \cite{garrido-alenda02}. Thus when more than one `word' needs to be output, they are joined with the \texttt{+} symbol, which is reserved for joining two `words' in one analysis, e.g. {\em өнүктүрөбүзбү} `TRANSLATION' is analysed as 
{\tt өнүк<v><tv><caus><aor><p1><pl>+бы<mod_qst>}

@@

In the case of Kyrgyz, dealing with the orthographical forms directly further simplifies some aspects of the morphophonology, since Kyrgyz orthography reflects a somewhat simplified version of the phonology: it ignores processes that the orthographies of many other Turkic do not, such as the phonemic distinction between velar and uvular consonants as well as sandhi voicing effects.

@@

\subsection{й+vowel letters}

In Kyrgyz, there are a series of letters which represent /й/ plus a vowel: /я, е, ё, ю/ represent /йа, йэ, йо, йу/ respectively.  However, /э/ is also represented as /е/ when short and after a consonant (i.e., it’s only represented as /э/ when long—/ээ/—or when word-initially).  These ``yoticised'' vowels proved difficult to work.  For example, /бой+(S)I/ \eng{length}, given normal vowel harmony rules, would surface as [бойу].  However, the correct form is [бою].  Distinct and separate vowel harmony rules had to be created for all vowel archiphonemes in post-vocalic contexts, e.g. here to turn \{I\} into [ю] after /й/.  An additional rule had to delete the underlying /й/ before yoticised vowel letters (so that e.g. [бойю] was not output).  Because there are too levels to the phonology in twolc and it’s otherwise not linear, attention had to be paid to the fact that e.g., the distinct vowel harmony rules for post-/й/ contexts recognised that the /й/ would not appear on the second [surface] level.

@@

\begin{table*}
	\centering
	\begin{tabular}{lr}
		\toprule
		Part of speech & Number of stems\\
		\midrule
		Noun & 4,972\\
		Verb & 1,231 \\
		Adjective & 944\\
		Proper noun & 796\\
		Adverb & 295 \\
		Numeral & 63 \\
		Conjunction & 58 \\
		Postposition & 51 \\
		Pronoun & 29\\
		Determiner & 27\\
		\midrule
		Total: 8,466 & \\
		\bottomrule
	\end{tabular}
	\caption{Breakdown of number of stems in the lexicon by major part of speech. Minor categories, such as partice, modal and copula are left out.}\label{table:stems}
	%FIXME: are we putting captions above or below tables?  I like them better below, but I got the impression that there's some standard to have them above.  Maybe it was just UW being table Nazis for my MA thesis.
\end{table*}

@@

\begin{table*}[htbp]
	\centering
	\begin{tabular}{lrrrr}
		\toprule
		Corpus           & Tokens    & Known     & Naïve coverage    & Mean ambiguity\\
		\midrule
		Kyrgyz Wikipedia & 329,524   & 270,668   & 82.11 $\pm$ 3.23  & 2.35\\
		RFERL Kyrgyzstan & 4,112,558 & 3614193   & 87.93 $\pm$ 1.25  & \\
		\bottomrule
	\end{tabular}
	\caption{FIXME}
	\label{table:cov}
\end{table*}

@@

The major work to be done is increasing the size of the lexicon. While a good level of coverage has been achieved with only 7,333 stems, real-word, or production morphological analysers have tens of thousands of stems, even for morphologically-rich languages like Kyrgyz.

@@

It is the hope of the authors that the work on this transducer will lead the way for work on transducers for other Turkic languages.  Indeed, transducers for Kazakh, Tatar, Bashqort, and Chuvash which are currently under development have benefited from work done on this transducer: many aspects have been based on the same general approach, and a number of phonological rules from the Kyrgyz transducer have served as the basis of rules in the transducers of these other languages.

>>>>

The analyser is designed to be used with the longest-match left-to-right (LRLM) tokenisation strategy, as described in \cite{garrido-alenda02}. Thus when more than one `word' needs to be output, they are joined with the \texttt{+} symbol, which is reserved for joining two `words' in one analysis, e.g. `өнүктүрөбүзбү ?' {\em TRANSLATION ?} is analysed as 
өнүк{\tt {\small <v><tv><caus><aor><p1><pl>+}}бы{\tt {\small <qst>}} two words: `өнүк' (verb, transitive, causative, aorist, first person, plural) and `бы' (question clitic).

@@

In the case of Kyrgyz, dealing with the orthographical forms directly further simplifies some aspects of the morphophonology, since Kyrgyz orthography reflects a somewhat simplified version of the phonology: it ignores processes that the orthographies of many other Turkic do not, such as the phonemic distinction between velar and uvular consonants as well as sandhi voicing effects. However, other aspects are made more complicated, such as й+vowel combinations (see §\ref{sec:yot}).

@@

\subsection{й+vowel letters}
\label{sec:yot}
In Kyrgyz, there are a series of letters which represent /й/ plus a vowel: /я, е, ё, ю/ represent /йа, йэ, йо, йу/ respectively.  However, /э/ is also represented as /е/ when short and after a consonant (i.e., it’s only represented as /э/ when long—/ээ/—or when word-initially).  These ``yoticised'' vowels proved difficult to work.  For example, /бой+(S)I/ \eng{length}, given normal vowel harmony rules, would surface as [бойу].  However, the correct form is [бою].  Distinct and separate vowel harmony rules had to be created for all vowel archiphonemes in post-vocalic contexts, e.g. here to turn \{I\} into [ю] after /й/.  An additional rule had to delete the underlying /й/ before yoticised vowel letters (so that e.g. [бойю] was not output).  Because there are too levels to the phonology in twolc and it’s otherwise not linear, attention had to be paid to the fact that e.g., the distinct vowel harmony rules for post-/й/ contexts recognised that the /й/ would not appear on the second [surface] level.

@@

\begin{table}
	\centering
	\begin{tabular}{lr}
		\toprule
		Part of speech & Number of stems\\
		\midrule
		Noun & 4,972\\
		Verb & 1,231 \\
		Adjective & 944\\
		Proper noun & 796\\
		Adverb & 295 \\
		Numeral & 63 \\
		Conjunction & 58 \\
		Postposition & 51 \\
		Pronoun & 29\\
		Determiner & 27\\
		\midrule
		Total: 8,466 & \\
		\bottomrule
	\end{tabular}
	\caption{Breakdown of number of stems in the lexicon by major part of speech. Minor categories, such as particle, modal and copula are left out.}\label{table:stems}
	%FIXME: are we putting captions above or below tables?  I like them better below, but I got the impression that there's some standard to have them above.  Maybe it was just UW being table Nazis for my MA thesis.
\end{table}

@@

\begin{table*}[htbp]
	\centering
	\begin{tabular}{lrrrr}
		\toprule
		Corpus           & Tokens    & Known       & Naïve coverage (\%) & Mean ambiguity\\
		\midrule 
		Kyrgyz Wikipedia & 329,524   & 270,668     & 82.1 $\pm$ 3.2      & 2.35\\
		RFERL Kyrgyzstan & 4,112,558 & 3,614,193   & 87.9 $\pm$ 1.2      & \\
		\bottomrule
	\end{tabular}
	\caption{Naïve coverage and mean ambiguity of the analyser over two test corpora.}
	\label{table:cov}
\end{table*}

@@

The major work to be done is increasing the size of the lexicon. While a good level of coverage has been achieved with only 8,466 stems, real-word, or production morphological analysers have tens of thousands of stems, even for morphologically-rich languages like Kyrgyz.

@@

It is the hope of the authors that the work on this transducer will lead the way for work on free/open-source tagset-compatible transducers for other Turkic languages.  Indeed, transducers for Kazakh, Tatar, Bashqort, and Chuvash which are currently under development have benefited from work done on this transducer: many aspects have been based on the same general approach, and a number of phonological rules from the Kyrgyz transducer have served as the basis of rules in the transducers of these other languages.

\section*{Acknowledgements}

We would like to thank: the Google Summer of Code 2011, which supported the development of the Turkish$\rightarrow$Kyrgyz MT system this transducer was designed for; Tolgonay Washington for native speaker intuitions of Kyrgyz and the anonymous reviewers for comments on how to improve the manuscript. This work has been partially funded by Spanish Ministerio de Ciencia e Innovación through project TIN2009-14009-C02-01.
